---
title: "Trade Dependence"
format: html
editor: source
---

Does trade dependence change states' foreign policy decisions? 

Policy-makers seek to manage their state's dependence on other states in a number of areas including economic and security spheres. One such area of concern is trade dependence. Literature exploring the effect of trade dependence on a state's foreign policy has produced conflicting findings.

A better understanding of dependence between states can be used to strengthen a state's political position in three ways. First, a state can better manage its dependence on other states and, thereby, strengthen its ability to resist other state's attempts at coercion. For example, a state can diversify its trade partner pool in an important product. Second, states can further develop other states' dependence on them, providing the state with more power to leverage others to act in its interest. For example, a state can encourage increased trading in another state's important products and build that state's dependence on it over time. Finally, states can anticipate and manage shifts in their neighbors', allies', and rivals' dependence on other states. For example, a state can anticipate whether a state in their region will acquiesce to demands from an increasingly dominant state.     

## Theory

Dependence on a foreign trade partner can alter the balance of power between two states. The dependent state needs to maintain access to the other state's market. The other state can leverage this dependence to compel the dependent state to act in the other state's interest (or not act against their interest). 

Outside of war, states rarely threaten to sever all trade ties with another state. However, states do frequently cease or slow down trade in a single or group of products as a means of inducing a change in their rival's behavior. Even the threat of changed trade flows can be used to compel a change in behavior. 

There are two central actors to this theory: the dependent state and the other state. Both consider the cost they face from severing or limiting trade in important products. The other state will strategically select products that cause more pain to the dependent state's economy than its own. To this end, it is important to distinguish between relationships characterized by asymmetric dependence and those characterized by mutual interdependence. A state's ability to credibly threaten to or use their stronger position in trade in a product important to another state is diminished when that state is dependent on trade in another product with the other country. 

The conditions under which changed trade flows can be used to leverage changes in another country's policies include: 

1.    The other country can find no alternative market. If the dependent state can find other countries to trade in their important product the material costs of severing trade are diminished. Where these costs are severely diminished, the threatening state is unlikely to be able to coerce policy concessions out of the threatened state.

1.    Most obviously, the target country cannot be able to retaliate with its own costly sanctions.

## Measuring trade dependence

I introduce a product-level measure of trade dependence. 

### Vital or important products

First, we need to discover which products are important to the country's economy. Trade in products that are vital to the economy should contract at a slower rate than the economy itself. If GDP is declining, we should see trade in products decline as well. The economy tries hard to ensure that it maintains access to these products, even when times are tough. 

```{r}
library(wbstats)
library(tidyverse)

# gdp <- wb_data("NY.GDP.MKTP.KN", country = "Australia", start_date = 2011, end_date = 2020, return_wide = F)

gdp_growth <- wb_data(
  "NY.GDP.MKTP.KD.ZG", 
  country = "Australia", 
  start_date = 2012, 
  end_date = 2021, 
  return_wide = F
) |> 
  arrange(date) |> 
  mutate(roc = value / lag(value) - 1)

ggplot(gdp_growth, aes(x = date, y = value)) + 
  geom_line() + 
  geom_hline(yintercept = 0) + 
  theme_minimal()
```

To explore: 
-   Number of years of importance
-   All positive roc, rather than simply a greater roc than GDP

```{r}
aus_trade <- rio::import(here::here("data-raw", "TradeData_1_27_2023_20_5_40.csv")) |> 
    janitor::clean_names()

trade <- aus_trade |> 
  arrange(cmd_code, period) |> 
  mutate(trade_growth = primary_value / lag(primary_value) - 1,
         trade_roc = trade_growth / lag(trade_growth) - 1) |> 
  left_join(
    gdp_growth,
    by = c("period" = "date")
  ) |> 
  select(period, cmd_code, cmd_desc, primary_value, trade_growth, value, roc, trade_roc) |> 
  mutate(score = trade_roc - roc,
         important = if_else(trade_roc > roc | trade_roc == roc, 1, 0))

aus_important <- trade |> 
  group_by(cmd_code) |> 
  summarise(important = sum(important, na.rm = T)) |> 
  ungroup() |> 
  filter(important > 5) |> 
  pull(cmd_code)
  
```


```{r}
aus_trade |> 
  filter(cmd_code %in% aus_important)
```

### Dependent state's potential costs

### Other state's potential costs

#### Cost of severing trade in that product

#### Dependent state's ability to retaliate

### Balance of costs
