---
title: "Calculating the HHI for International Trade Markets"
author: "Harriet Goers"
format: 
  html:
    self-contained: true
editor: visual
execute:
  message: false
  warning: false
---

```{r}
library(tidyverse)
library(rio)
library(here)
library(janitor)
```

First we need to read in our total trade data:

```{r}
df_raw <- map_dfr(
  list.files(here("data-raw", "total_trade"), full.names = T), 
  ~ read_csv(.x, col_types = "ccccccccccccccccccccccccccccccccccc")
)
```

Next we need to calculate the HHI for each product:

```{r}
df_hhi <- df_raw |> 
  clean_names() |> 
  select(year = yr, direction = rg_desc, reporter = rt_title, partner = pt_title, cmd_code, trade_value) |> 
  group_by(cmd_code, year, direction) |> 
  mutate(
    year = as.numeric(year),
    trade_value = as.numeric(trade_value),
    global_trade_value = sum(trade_value),
    reporter_share = trade_value / global_trade_value * 100,
    hhi = sum(reporter_share^2)
  ) |> 
  ungroup() |> 
  left_join(
    read_csv(here("data", "product_codes.csv"), col_types = c(id = "c")) |> 
      select(cmd_code = id, cmd_desc = text)
  ) |> 
  relocate(cmd_desc, .after = cmd_code)

df_hhi
```

Now we can identify which markets are highly concentrated:

```{r}
df_hhi |> 
  filter(hhi > 2500) |> 
  distinct(year, direction, cmd_code, cmd_desc, hhi) |> 
  arrange(desc(hhi))
```

And we can identify the dominant traders in each market. Let's look at 010612 - Mammals; live, whales, dolphins and porpoises in 2019, which is a highly concentrated market.

```{r}
df_010612 <- filter(df_hhi, cmd_code == "010612")

arrange(df_010612, desc(reporter_share))
```

```{r}
ggplot(df_010612, aes(x = reporter_share, y = reorder(reporter, reporter_share))) + 
  geom_col() + 
  theme_minimal() + 
  labs(x = "Reporter share (%)", y = NULL)
```

I reference the full range of HHI in my article, so I am going to save a summary of each product-year HHI calculated.

```{r}
df_hhi |> 
  distinct(year, direction, cmd_code, cmd_desc, hhi) |> 
  export(here("data", "hhi_summary.csv"))
```

I analyse highly concentrated markets. To do this, I need data on the reporter share in highly concentrated markets.

```{r}
df_hhi |> 
  filter(hhi > 2500) |> 
  export(here("data", "hc_product_global.csv"))
```

Finally, I create a summary that lists only the top trader in each of these highly concentrated markets:

```{r}
df_hhi |> 
  filter(hhi > 2500) |> 
  group_by(year, direction, cmd_code) |> 
  slice_max(reporter_share) |> 
  export(here("data", "hc_product_global_top_reporter.csv"))
```
