---
title: "Calculating the HHI for International Trade Markets"
author: "Harriet Goers"
format: 
  html:
    self-contained: true
editor: visual
execute:
  message: false
  warning: false
---

```{r}
library(tidyverse)
library(rio)
library(here)
library(janitor)
```

First we need to read in our total trade data:

```{r}
df_raw <- map_dfr(
  list.files(here("data-raw", "2019"), full.names = T), 
  ~ read_csv(.x, col_types = "ccccccccccccccccccccccccccccccccccc")
)
```

Next we need to calculate the HHI for each product: 

```{r}
df_hhi <- df_raw |> 
  clean_names() |> 
  select(year = yr, direction = rg_desc, reporter = rt_title, partner = pt_title, cmd_code, cmd_desc = cmd_desc_e, trade_value) |> 
  group_by(cmd_code, year) |> 
  mutate(
    year = as.numeric(year),
    trade_value = as.numeric(trade_value),
    global_trade_value = sum(trade_value),
    reporter_share = trade_value / global_trade_value * 100,
    hhi = sum(reporter_share^2)
  ) |> 
  ungroup()

df_hhi
```

Now we can identify which markets are highly concentrated: 

```{r}
df_hhi |> 
  filter(hhi > 2500) |> 
  distinct(year, cmd_code, cmd_desc, hhi) |> 
  arrange(desc(hhi))
```

And we can identify the dominant traders in each market. Let's look at 0502: Pigs', hogs' or boars' bristles and hair; and waste thereof in 2019, which is a highly concentrated market. 

```{r}
df_0502 <- filter(df_hhi, cmd_code == "0502")

arrange(df_0502, desc(reporter_share))
```

```{r}
ggplot(df_0502, aes(x = reporter_share, y = reorder(reporter, reporter_share))) + 
  geom_col() + 
  theme_minimal() + 
  labs(x = "Reporter share (%)", y = NULL)
```

I am interested in highly concentrated markets, so I am going to use these summaries going forward. 

```{r}
df_hhi |> 
  filter(hhi > 2500) |> 
  export(here("data", "hc_product_level_trade.csv"))
```
