---
title: "A New Measure of Trade Dependence"
format: 
  html:
    self-contained: true
    toc: true
editor: visual
bibliography: references.bib
execute: 
  echo: false
  message: false
  warning: false
---

```{r libraries}
library(tidyverse)
library(scales)
library(here)
library(tidygraph)
library(ggraph)
```

```{r data}
product_codes <- read_csv(here("data", "product_codes.csv"))
hhi_summary <- read_csv(here("data", "hhi_summary.csv"), col_types = c(cmd_code = "c"))
hc_summary <- read_csv(here("data", "hc_product_global.csv"))
hc_summary_top_reporter <- read_csv(here("data", "hc_product_global_top_reporter.csv"), col_types = c(cmd_code = "c"))
hv_summary <- read_csv(here("data", "highly_vulnernable_bilat_trade_summary.csv"))
hv_trade <- read_csv(here("data", "highly_vulnernable_bilat_trade.csv"))
```

Are states able to exploit trade imbalances to coerce other states into changing their behavior?

I adopt @carnegie2015's definition of coercion: "the use of threats and assurances in combination to influence another state's behavior".

Systematic exploration of trade dependence requires a comprehensive and detailed picture of the whole network of interactions between states. If a state is going to use trade dependence to compel another state to act in its interest, it needs to consider how it maximizes the economic cost it can threaten to impose on the target whilst taking into account how the target and other trading states will react to that threat. A state can only prompt this change in behavior if the economic cost of the threatened trade disruption is strictly greater than the cost the target will bear from changing their behavior. When these conditions hold, the target state is trade dependent on the imposing state.

We need to understand both the economic cost the target faces from the trade ban, and the costs it expects to bear from acting in the imposing state's interests instead of its own. This paper will provide a novel commodity-level measure of the expected economic cost of trade disruptions. This measure captures the strategic calculations made by three important actors: the imposing state, the target state, and other trading states participating in the market for a commodity. I will then use this measure of trade dependence to explore the efficacy of this foreign policy tool.

Which states control dominant shares of trade in critical products? How can target states attempt to fill a potential or existing gap in their access to these products or markets? How can other states leverage these gaps?

This novel measure of trade dependence is generic: it can be used to explore power dynamics between any pairing of states. It allows scholars and policymakers to systematically capture the potential for a dominant state to pressure a trade partner to act in the dominant state's interest.

This paper makes four contributions. First, it builds a novel measure of trade dependence that captures the strategic interactions exhibited by states as they interact in global markets. Second, it identifies cross-sectional and temporal patterns in trade dependence. Third, it explores the relationship between dependence and X: a common and long-standing debate in international relations literature. Finally, it explores the implications and potential applications of this analysis for ongoing policy challenges and scholarly debate.

## Literature review

The link between dependence and power has long been chewed over. Broadly, the literature has debated whether increasing economic dependence on a state increases the likelihood that the partner will accommodate the reporter's interests, *ceteris paribus*.

Previous research has relied on gross measures of trade dependence, including total trade and total trade as a proportion of a country's gross domestic product or gross national income. This is despite the fact that we rarely see imposers of trade disruptions attempt to disrupt all trade with a partner. Rather, they strategically select specific products or sectors to target [@gholz2021]. States are surprisingly resilient to trade bans or disruptions. Our surprise should prompt us to re-examine how we understand and measure trade dependence.

The majority of scholars look at the overall trade relationship between two states [@barbieri1996].

Several scholars have sought to expand on this research by looking at commodity-level trade relationships [@chatagnier2017].

Several scholars have included in their measure trade with a potential target's allies. However, these measures of dependence also examine the overall trade relationship between the potential instigator and these allied states [@chen2021].

Several researchers have found evidence that increased trade with China has led to increased convergence on Beijing's position on issues [@flores-macías2013; @strüver2014; @kastner2016]. Kastner finds stronger support for the claim that increased trade leads to increased convergence with China on economic issues. Specifically, he finds a strong relationship between increased trade and support for China's position on its status as a market economy.

Others have found evidence to the contrary, including @medeiros2008 and @ross2006.

## Theory

States choose to develop trade relationships with other states. Not only do they seek to take advantage of the benefits of specialization and scale, but they seek to manage the risks attached to delegating development and production in certain good. Included in this risk is the possibility that the relationship with the trade partner will deteriorate to the point where it will seek to leverage this vulnerability to extract a costly concession from the state [@carnegie2015]. If the imposing state can credibly threaten to impose sufficiently large costs on the target state that make the concession the utility-maximizing choice, the target state is trade dependent on the imposing state. This analysis starts after the target state has decided to develop a trade relationship with the imposing state. It is important to acknowledge that some states will select out of developing a trade relationship with certain states in the first place. This is particularly the case when those states believe that the risk that the imposing state will leverage its dominant trade position to extract policy concessions from the target is high. We will explore the implications of this process for trade dependence in the formal model to follow.

### A new definition of trade dependence

To determine whether a target state is vulnerable to coercive trade policies by another state, we need to understand how that target state can react to the gap created by trade disruptions. Can they fill the gap, either by locating alternate trade partners or by increasing domestic production or consumption of a commodity? Alternatively, can they adapt to living with the gap by finding appropriate substitutions for the commodity or reducing their demand for the commodity and any downstream products that rely on access to that commodity?

The imposing state faces potentially considerable material and reputational costs from threatening to, or directly imposing, coercive trade policies against a target state. This is the case whether or not the target state is trade dependent on the imposing state. For example, if other states believe that there is a significant risk that the imposing state will leverage its dominant market share to coerce costly foreign policy concessions from them, they will not enter into a trade relationship with the potential imposing state [@carnegie2015]. If these states observe the imposing state using its leverage against a target state, they will increase their beliefs about this risk. They are, therefore, more likely to forego developing a trade relationship with the imposer. The imposing state, therefore, misses out on the material benefits of a more efficient and diverse trade partner pool. This analysis focuses on dynamics between the target and imposer state. Therefore, we start after the target state has decided to develop a trade relationship with the imposer state. We can, therefore, understand this cost as a component of the imposing state's expected costs from threatening or initiating coercive trade policies against the target, not as a component of the target state's decision calculus.

This dynamic does not depend on the target state's ability to resist the imposing state's coercion. However, if the target state is trade dependent on the imposing state, the imposing state gains the concession from the target. Where these concessions derive greater benefits for the imposing state than the costs of foregone trade with the target state and other current or potential trade partners, it has an incentive to use its dominant trade position to extract these concessions. If the target state is not dependent on the imposing state because it can either fill or adapt to living with the gap left by the imposing state, the imposing state only incurs the costs of this coercive policy without deriving any benefits. This is why is it important to consider all three important actors involved in this process: the imposing state, the target state, and alternative trade partners.

It is also important to explore the strategic calculations made by each of these actors. First, let us examine the imposing state's options. If the imposing state is in a position to develop or maintain a dominant trade position in a market, coercive trade policies sit in its foreign policy tool kit. Does it use this tool? The imposing state expects to incur the costs outlined above if it chooses to do so. It will, therefore, only impose coercive trade policies against a target state when it expects to reap greater or equal benefits from the concession extracted from the target than this expected cost. This prompts two important questions: when can the imposing state expect the concession to be extracted, and what is the net benefit it expects to derive?

When can the imposing state expect to extract the desired concession from the target state? I argue this occurs when the target state is trade dependent on the imposing state. For this to be the case, two conditions must hold. First, the target state must be unable to fill the gap left by the target state. This can be the result of the target being unable to locate alternate trade partners, or the target state being unable to fill the gap through increased domestic production or consumption. Second, the target state cannot adapt to living with the gap left by the imposing state. This can be the result of no viable substitutions being available for the commodity in question, or the target state's economy being unable to operate without the commodity or the downstream products which use the commodity as an input. Where these conditions hold, the imposing state should expect to extract the desired concession from the target state.

Why would a target state put itself into such a vulnerable position in the first place? States are concerned with and attempt to manage their level of vulnerability to others' coercive trade policies [@carnegie2015]. The target state decides whether to invest in and sustain a noncompetitive trade relationship with another state. The target state understands that a shock could occur with some probability that sours their relationship with their dominant trade partner and prompts the partner to threaten to or directly impose coercive trade policies. The target state, as a strategic actor, will only enter into this relationship when the benefits they gain from access to the commodity are sufficiently high, or their beliefs about the risk of the dominant trade partner exploiting their position to extract concessions from the target are sufficiently low.

Let us return to the imposing state. When will the imposing state expect the benefits of imposing a coercive trade policy on a trade dependent target to outweigh the costs it will incur? This depends on the outcome they seek to achieve [@dahl1983; @baldwin2020]. To answer this question, we need to specify what concession the imposing state hopes to extract and the level of importance that concession has to the imposing state. I will expand on this in my empirical analysis later in this article.

This measure of trade dependence directly addresses the three questions prompted by the process outlined above. First, we ask whether the imposing state can credibly threaten to impose severe economic costs on the target. Does their threat have teeth? As @gholz2021 demonstrate, to be able to credibly threaten to or directly impose a significant economic cost on a target, the imposing state needs to have market power. To illustrate, let us look at export trade dependence, or the target state's level of vulnerability to severance of exports in commodities from the imposing state. At the extreme, if the imposing state is the only supplier of a commodity, the target state will not be able to locate alternative suppliers. It will need to ramp up domestic production, find a substitute, or reduce its consumption of the commodity and any downstream products. Alternatively, if the imposing state is one of many suppliers of the commodity, the target state should be able to locate an alternative supplier and only suffer slightly increased prices for the product on global markets.

If the imposing state has sufficient market power to credibly threaten to stifle the target state's access to a commodity on the global market, we can move on to asking how the target and other states can respond to this threat. Broadly, the target state has two options. First, they can locate alternative suppliers, or increase domestic production, of the commodity and attempt to maintain their current levels of consumption. Second, they can attempt to reduce their consumption of the product through contractions or substitutions. In short, they can attempt to fill the gap or adapt to living with it.

Let us first look at how the target state can attempt to maintain current levels of access to, or consumers of, the commodity. To fill the gap, the target can attempt to locate alternative trade partners. The imposing state cannot threaten to impose large economic costs on the target of trade restrictions if the target can easily locate alternative suppliers of the product [@peksen2016; @gholz2021]. We, therefore, need to understand how other states with the capacity to trade in the threatened commodity will respond to this threat. For simplicity, we will assume that all other states that trade in the commodity will look to fill the gap. Exporters of the commodity will be attracted to the higher price at which they can sell the commodity to the target. Importers of the commodity will demand it at a lower price than they previously accessed.

> TO DO: Explore why this is silly. Previous work has demonstrated the importance of alliances in shaping this response by other states [@chen2021]. Perhaps this is better in a different paper.

The target state may also be able to ramp up or establish domestic production of the commodity to fill the gap left by the imposing state. This option is not always viable, particularly for natural resources or highly specialized commodities.

Alternatively, the target state can adapt to living with the gap. One option available to the target state is to locate viable substitutions for the threatened commodity. We can model how the target state would increase its access to these substitutes similarly to how we measure their expected access to the threaten commodity.

The target state can otherwise attempt to reduce its demand for that commodity and any downstream products that use the commodity as a critical input. This would result in an overall reduction in the amount the target state consumes of, or generates revenue from, a commodity.

If the target state cannot fill the gap or adapt to living with the gap, it is trade dependent on the imposing state. We should expect to be able to observe this dependence through greater policy convergence [@flores-macías2013; @strüver2014] or direct coercion [@kastner2016]. In short, the target state is vulnerable to the imposing state's coercive trade policy.

In summary, whether or not the target state is trade dependent on an imposing state is a product of the target's ability to fill the gap created or threatened by the imposing state. This is intuitive, but it often poorly measured in international relations scholarship.

### Formal model of coercive trade

The model consists of interactions between the imposing state (I) and the target state (T) and the target state and alternative trade partners (A).[^1] We will model both export and import trade dependence. Let us start with import trade dependence.

[^1]: An interesting extension to this model would be to incorporate some understanding of the relationship between the imposing state (I) and the alternative trade partners (A). Why did some potential trade partners elect not to trade with Russia following the imposition of Ukraine war-related sanctions? Is this more important than Russia's relationship with the trade partners that increased their trade relationship with it?

The target state produces a good that is demanded by the imposing state. The target state produces this good with some competitive advantage: the imposing state is better off importing the good from the target state than developing or maintaining domestic production of the good. The type of good is irrelevant. What is important is that the imposing state demands this good and that the target state is more efficient at producing it than the imposing state.

The amount of the good exported by the target state to the imposing state is determined by the tariff, $\tau$, adopted by the imposing state following negotiations with the target state. The imposing state can set any tariff rate, $\tau \epsilon [0, \bar{\tau})$, where $\bar{\tau}$ is the rate for which it is no longer more efficient for the target state to export the good to the imposing state. For example, if the target state can produce a good and sell it in the imposing state for one hundred dollars and the imposing state's domestic market can produce and sell the good for one hundred and fifty dollars, the *ad valorem* tariff can be set anywhere between $\tau \epsilon [0, 0.5)$. Any tariff rate higher than 50 percent means that the imposing state's domestic market will only be able to consume domestically-produced goods at a cost of one hundred and fifty dollars and the target state does not export to the imposing state. Where $\tau = \bar{\tau}$, the imposing state is indifferent between domestically producing or importing the good.

It follows that the amount of the good available for export and consumption is impacted by the tariff. Related, we need to understand how tolerant the imposing state's economy is to price increases. If a small increase in prices leads to a steep decrease in demand for the product, the imposing state has greater control over flows of the good into its economy than if demand for the good is relatively inelastic. The imposing state's type is represented by $\zeta_I = \{L, H\}$, where $\zeta_I = L$ indicates that the imposing state's economy has a low level of tolerance to price increases and $\zeta_I = H$ indicates that the imposing state's economy has a high level of tolerance to price increases.

Therefore, I model the target and imposing states' utility over the good as functions of the tariff and the imposing state's type: $\mu_i(\tau, \zeta_I)$, where $i = \{I, T\}$. The target's utility over the good, $\mu_T(\tau, \zeta_I)$, is assumed to be concave in $\tau$ and monotonically decreasing in $\tau$.

The imposing state's utility, $\mu_I(\tau, \zeta_I)$ is assumed to have an interior maximum at $\bar{\tau}$. The imposing state has to work within the constraints of balancing the benefits of a higher tariff, including potential higher tax revenues, with providing its economy with access to an in-demand good at a cheaper price than could otherwise be accessed.

The target and imposing states start the game with a tariff rate that is set through cooperative agreement. This process is not modeled in this analysis.

The target state regularly invests in the industry to allow it to produce sufficient goods to export to the imposing state. For example, the target state invests in building and maintaining new factories that expand the quantity of the good it can produce, allowing it to meet the imposing state's demands. This cost is represented by $c_T = \{0, \bar{c_T}\}$, where $c_T = \bar{c_T}$ when the target state injects sufficient investment in the industry to meet the imposing state's demands. $c_T = 0$ when the target state decides against incurring this cost and reduces its ability to export competitively to the imposing state. Importantly, the target state enjoys a competitive advantage in producing this good relative to the imposing state. Therefore, we can assume that $\bar{c_T} < \bar{c_I}$, where $\bar{c_I}$ represents the cost the imposing state would need to incur to establish domestic production capacities that can produce the good equally efficiently to the target state. The imposing state's decision to invest in the industry producing the good is represented by $c_I = \{0, \bar{c_I}\}$. When the imposing state imports the good from the target state $c_I = 0$.

The game starts with a disagreement between the target and imposing states. The imposing state requests that the target state makes some concession. The target state then decides whether to make the concession or reject the request: $f = \{0,1\}$, where $f = 1$ indicates that the target state conceded. Making this concession is costly for the target state. This cost is represented by $x_C$.

The imposing state can use a myriad of coercive foreign policy tools to compel the target state to concede. These include economic, diplomatic, and military options. For simplicity, my model includes two options: coercive trade policies and other coercive policies. The utility gained by this other relationship is modeled as $\mu_i(\alpha)$, where $i = \{I, T\}$.

Focusing on coercive trade policies, the imposing state can change the tariff rate applied to the good exported from the target state to attempt to compel the target state to concede. The imposing state can threaten to increase the tariff rate and, thereby, decrease the profits derived by the target state from its exports to the imposing state. Alternatively, the imposing state can offer to lower the tariff rate in exchange for the policy concession. This carrot increases the profits the target can make from exporting the good to the imposing state.

The imposing state is attempting to impose on the target state sufficient costs to overcome the cost to the target state of conceding, $x_C$.

In summary, the game proceeds as follows:

1.  The target and imposing state establish a trade relationship in relation to a specific good with a tariff rate of $\tau^*$.

2.  A disagreement occurs.

3.  The imposing state threatens to change the rate of $\tau$, $\alpha$, or a combination of both to compel the target state to concede, $f = 1$.

4.  The target state chooses $f = \{0,1\}$.

5.  The imposing state sets a new $\tau$ and/or $\alpha$.

6.  The target state looks to mitigate the cost of this change in $\tau$ and/or $\alpha$, including by seeking alternative trade partners, increasing domestic consumption of the good, or adapting to reduced revenue from exports.

7.  The target and imposing states hold out until the expected cost of resisting is greater than (or equal to) the expected cost of conceding.

## Empirics

### Identifying trade dependence

Can the imposing state credibly threaten to impose severe economic costs on the target state? To answer this question, we need to measure trade market concentration. At the extreme, if the imposing state is the only supplier of a commodity, the target state will not be able to locate alternative suppliers. It will either need to find a substitute for, or reduce its consumption of, the commodity and any downstream products. Alternatively, if the imposing state is one of many suppliers of the commodity, the target state should be able to locate an alternative supplier and only suffer slightly increased prices for the commodity on the global markets.

The Herfindahl-Hirschman Index (HHI) is a widely used measure of market concentration. When applied to international trade markets, the HHI measures the value of a commodity traded by each state relative to the total value of trade in the commodity. It is calculated as the sum of the square of each state's market share. It ranges from zero to 10,000, with values nearing zero indicating a highly competitive market and values nearing 10,000 indicating a highly concentrated or monopolized or monopsonized market.

This analysis uses the HHI to identify commodity trade markets in which the dominant participant could credibly threaten to stem a target state's international access to the commodity. I examine market concentration for all commodities traded globally across our period of interest. This includes `r nrow(product_codes) |> comma()` different commodities.[^2]

[^2]: This analysis defines commodity groupings according to the six digit Harmonized Tariff Schedule (HS). The HS is a standardized classification system for traded products. It is used globally to identify products when collecting taxes, assessing duties, and reporting trade statistics. A full list of commodities and their HS codes can be accessed from [UN COMTRADE](https://comtrade.un.org/Data/cache/classificationHS.json).

Using trade data provided by the UN COMTRADE, I find that across all `r nrow(product_codes) |> comma()` different commodities the HHI ranges from `r hhi_summary |> summarise(min(hhi)) |> round()` to `r hhi_summary |> summarise(max(hhi)) |> round() |> pull() |> comma()` across our period of interest. Commonly, a market with a HHI of more than 2,500 is considered to be highly concentrated. `r hhi_summary |> count(hc = hhi > 2500) |> mutate(prop = n / sum(n)) |> filter(hc == T) |> pull(prop) |> percent()` of commodities are traded in highly concentrated markets for at least one year in our period of interest. These highly concentrated markets represent the trade in commodities for which a state or a small set of states can potentially restrict a target state's international access. The dominant trading states in these markets represent potential viable imposing states. Their trade partners represent potential target states.

```{r}
ggplot(hhi_summary, aes(x = hhi)) + 
  geom_histogram() +
  geom_vline(xintercept = 2500) + 
  theme_bw() + 
  labs(
    x = "HHI",
    y = NULL
  ) + 
  scale_x_continuous(limits = c(0, 10000))
```

We can now examine the target state's response to a potential trade disruption of a vulnerable commodity. First we will determine whether the target state can fill the gap created by the trade ban. The target can either look to other trading states to fill this gap, or it can increase domestic production in the commodity.

```{r}
alt_reporter_null <- filter(hc_summary_top_reporter, reporter_share == 100)

alt_reporter_factor <- hc_summary_top_reporter |> 
  anti_join(alt_reporter_null, by = c("year", "direction", "reporter", "partner", "cmd_code", "cmd_desc", "trade_value", "global_trade_value", "reporter_share", "hhi")) |> 
  filter(reporter_share > 50) |> 
  select(reporter, cmd_code, reporter_share) |> 
  mutate(alt_reporter_share = 100 - reporter_share, factor = round(reporter_share / alt_reporter_share, 3))
```

Can the target state locate alternative trade partners to fill the gap left by the imposing state? To measure this, I use annual bilateral trade data describing trade between the most dominant trading state and all of their trade partners for all commodities identified to be traded in highly concentrated markets. Attracted by the higher prices exports would receive or lower prices at which imports could be accessed, could alternative trade partners divert their trade in the vulnerable commodity to the target and fill the gap?[^3] For `r hc_summary_top_reporter |> filter(reporter_share > 50) |> nrow() |> comma()` dyad-commodity pairings, the potential imposing state dominates trade to such an extent that all alternative trade partners combined cannot fill the gap they could leave. In fact, in `r nrow(alt_reporter_null) |> comma()` commodity-years, a state is the sole trader in a commodity. For commodities that are traded by at least two states, all other trading states combined would need to increase production by a factor of at least `r alt_reporter_factor |> summarise(min(factor))` and up to `r alt_reporter_factor |> summarise(max(factor)) |> pull() |> comma()` to fill the gap left by the imposing state. A full list of these products and dominant trade partners is provided in the appendix. For the remaining dyad-commodity pairings, trade diverted from some combination of alternative trade partners could fill the gap left by the imposing state.

[^3]: This is a very strong test of dependence. It is highly unlikely that any single alternate trade partner would respond to a trade ban by diverting all of its trade in a commodity from existing trade partners to the target state. In many cases, this action is impossible. My test is even less realistic: I ask whether all alternate trading states would divert all of their trade to the target state. I am comfortable with this test because it makes it harder for me to prove significance. At later stages in this process, I will reduce this threshold from 100 percent of trade diverted to a lower threshold.

Can the target state increase domestic production to fill the gap? As a rather blunt measure of this, I will assume that the target state will divert all current exports of the commodity to their domestic market. Although the target state in `r hv_trade |> filter(partner == alt_reporter) |> nrow()` of these highly vulnerable markets exports the vulnerable commodity, even if they divert these exports to their domestic market they will remain unable to fill the gap left by the imposing state.

Next, we need to consider the cost borne by the imposing state. There are two factors that I include in this analysis: the economic cost of imposing a trade ban, and the potential for retaliation from the target state. The economic cost borne by the imposing state includes the direct loss of revenue from trade in the product at the time the trade ban is imposed, and expected losses from future trade in the product for the duration of the ban. I use bilateral trade data at the commodity level to calculate the former component of this measure. To determine the expected cost of the trade ban in the medium- to long-term, we need to understand the imposing state's beliefs about how resilient the target state will be to this trade disruption. If the imposing state believes that the target state will cave to its demands quickly, the expected future cost of trade disruptions will be smaller than if the imposing state believes that the target state will resist.

> TO DO: Test different discount factors. Can I simulate the range of beliefs that each potential imposing state would need to impose the trade ban? Also should probably include a general future discount.

```{r}
inter_df <- hv_summary |> 
  distinct(r = reporter, p = partner) |> 
  bind_rows(hv_summary |> distinct(p = reporter, r = partner)) |> 
  count(p, r) |> 
  filter(n > 1)
```

The imposing state could also expect that the target state will react to trade disruptions by imposing a retaliatory trade ban. To identify the imposing states for which this could serve as a deterrent, I identify interdependent dyads. Interdependence is rare: only `r (nrow(inter_df) / 2) |> broman::spell_out()` dyads exhibit it. Each state in these pairings could significantly limit their opposing state from accessing a commodity in retaliation to a threatened or imposed trade ban. For critical goods, this should serve as a sufficient deterrent to either state imposing a trade ban.

```{r}
reporters <- distinct(hv_summary, entity = reporter)
partners <- distinct(hv_summary, entity = partner)

nodes <- reporters |> 
  bind_rows(partners) |> 
  distinct()

edges <- hv_summary |> 
  select(reporter, partner, cmd_desc)

dep_network <- tbl_graph(
  nodes = nodes, 
  edges = edges,
  directed = T
)

ggraph(dep_network, layout = "fr") +
  geom_edge_link(arrow = arrow(length = unit(2, 'mm')), alpha = 0.5) + 
  geom_node_point() + 
  geom_node_text(aes(label = entity), repel = TRUE) + 
  theme_graph()
```

> TO DO: Look for circles. Next paper?

### Controls

Overall bilateral trade flows [@barbieri1996].

Overall bilateral import and export flows [@kastner2016; @strüver2014].

Membership in the World Trade Organisation [@carnegie2015].

## Measuring influence

We first need an appropriate measure of political influence. It is difficult to operationalize political influence across countries and time [@kastner2016].

Both Flores-Macias and Kreps [-@flores-macías2013] and Struver [-@strüver2014] measure the distance between a country's and China's voting pattern in the United Nations General Assembly. Kastner [-@kastner2016] examines countries' positions on issues pertaining to issues over which China cares deeply: Taiwan and Tibet, and its status as a market economy.

## Appendix

**Complete list of dominated pairings:**

```{r}
hc_summary_top_reporter |> 
  filter(reporter_share > 50) |> 
  transmute(year, direction, reporter, commodity = cmd_desc, reporter_share = percent(reporter_share / 100, accuracy = 0.01)) |> 
  arrange(reporter, year) |> 
  janitor::clean_names("title") |>
  knitr::kable()
```
