---
title: "A New Measure of Trade Dependence"
format: 
  html:
    self-contained: true
    toc: true
editor: visual
bibliography: references.bib
execute: 
  echo: false
  message: false
  warning: false
---

```{r libraries}
library(tidyverse)
library(scales)
library(rio)
library(here)
```

```{r data}
product_codes <- import(here("data", "product_codes.csv"))
hhi_summary <- import(here("data", "hhi_summary.csv"))
hc_summary <- import(here("data", "hc_product_global.csv"))
hv_summary <- import(here("data", "highly_vulnernable_bilat_trade.csv"))
```

Are states able to exploit trade imbalances to coerce other states into changing their behavior?

Systematic exploration of trade dependence requires a comprehensive and detailed picture of the whole network of interactions between states. If a state is going to use trade dependence to compel another state to act in its interest, it needs to consider how it maximizes the economic cost it can threaten to impose on the target whilst taking into account how the target and other trading states will react to that threat. A state can only prompt this change in behavior if the economic cost of the threatened trade disruption is strictly greater than the cost the target will bear from changing their behavior. When these conditions hold, the target state is trade dependent on the imposing state.

We need to understand both the economic cost the target faces from the trade ban, and the costs it face from acting in the imposing state's interests instead of its own. This paper will provide a novel commodity-level measure of the expected economic cost of trade disruptions. This measure better reflects the strategic calculations made by three important actors: the imposing state, the target state, and other trading states participating in the market for a commodity. It will then use this measure of trade dependence to explore the efficacy of this foreign policy tool.

Which states control dominant shares of trade in critical products? How can target states attempt to fill a potential or existing gap in their access to these products or markets? How can other states leverage these gaps?

This novel measure of trade dependence is generic: it can be used to explore the power dynamic between any pairing of states. It allows scholars and policymakers to systematically capture the potential for a dominant state to pressure a trade partner to act in the dominant state's interest.

This paper makes four contributions. First, it builds a novel measure of trade dependence that better captures the strategic interactions exhibited by states as they interact in global markets. Second, it identifies cross-sectional and temporal patterns in dependence across trade. Third, it explores the relationship between dependence and X: a common and long-standing debate in international relations literature. Finally, it explores the implications and potential applications of this analysis of dependence for ongoing policy challenges and scholarly debate.

## Literature review

The link between dependence and power has long been chewed over. Broadly, the literature has debated whether increasing economic dependence on a state increases the likelihood that the partner will accommodate the reporter's interests, ceteris paribus.

Previous research has relied on gross measures of trade dependence, including total trade and total trade as a proportion of a country's gross domestic product or gross national income. This is despite the fact that we rarely see imposers of trade disruptions attempt to disrupt all trade with a partner. Rather, they strategically select specific products or sectors to target [@gholz2021]. States are surprisingly resilient to trade bans or disruptions. Our surprise should prompt us to re-examine how we understand and measure trade dependence.

The majority of scholars look at the overall trade relationship between two states [@barbieri1996].

Several scholars have sought to expand on this research by looking at commodity-level trade relationships [@chatagnier2017].

Several scholars have included in their measure trade with a potential target's allies. However, these measures of dependence also examine the overall trade relationship between the potential instigator and these allied states [@chen2021].

Previous research on weaponized dependence

Several researchers have found evidence that increased trade with China has led to increased convergence on Beijing's position on issues [@flores-macías2013; @strüver2014; @kastner2016]. Kastner finds stronger support for the claim that increased trade leads to increased convergence with China on economic issues. Specifically, he finds a strong relationship between increased trade and support for China's position on its status as a market economy.

Others have found evidence to the contrary, including @medeiros2008 and @ross2006.

## Theory

### A new definition of trade dependence

To determine whether a target state is vulnerable to coercive trade policies by another state, we need to understand how that target state can react to the gap created by trade disruptions. Can they fill the gap, either by locating alternate suppliers or by increasing domestic production of a commodity? Can they adapt to living with the gap by finding appropriate substitutions for the commodity or reducing their demand for the commodity and any downstream products that rely on access to that commodity?

This measure of trade dependence directly address the three questions prompted by the process put forward above. First, we ask whether the imposing state can credibly threaten to impose severe economic costs on the target. Does their threat have teeth? As @gholz2021 demonstrate, to be able to credibly threaten to, or directly, impose a significant economic cost on a target, the imposing state needs to have market power. For simplicity, we will focus on export trade dependence, or the target state's vulnerability to severance of exports in commodities from the imposing state. At the extreme, if the imposing state is the only supplier of a commodity, the target state will not be able to locate alternative suppliers. It will either need to find a substitute or reduce its consumption of the commodity and any downstream products. Alternatively, if the imposing state is one of many suppliers of the commodity, the target state should be able to locate an alternative supplier and only suffer slightly increased prices for the product on global markets.

If the imposing state has sufficient market power to credibly threaten to stifle the target state's access to a commodity on the global market, we can move on to asking how the target and other states can respond to this threat. Broadly, the target state has two options. First, they can locate alternative suppliers of the commodity and attempt to maintain their current levels of consumption. Second, they can attempt to reduce their consumption of the product through contractions or substitutions. In short, they can attempt to fill the gap or adapt to living with it.

Let us first look at how the target state can attempt to maintain current levels of access to, or consumers of, the commodity. To fill the gap, the target can attempt to locate alternative trade partners. The imposing state cannot threaten to impose large economic costs on the target of trade restrictions if the target can easily locate alternative suppliers of the product [@peksen2016; @gholz2021]. We, therefore, need to understand how other states with the capacity to trade in the threatened commodity will respond to this threat. For simplicity, we will assume that all other states that trade in the commodity will look to fill the gap. Exporters of the commodity will sell to the target at a higher price and importers of the commodity will demand it at a lower price than they previously accessed.

> TO DO: Explore why this is silly. Previous work has demonstrated the importance of alliances in shaping this response by other states [@chen2021].

> H1a: A target state is increasingly vulnerable to export restrictions with the imposing state's export market concentration, *ceteris paribus*.
>
> H1b: A target state is increasingly vulnerable to import restrictions with the imposing state's import market concentration, *ceteris paribus*.

The target state may also be able to ramp up or establish domestic production of the commodity to fill the gap left by the imposing state. This option is not always viable, particularly for natural resources or highly specialized commodities.

> H2a: A target state is less vulnerable to export restrictions when they have domestic production capabilities.
>
> H2b: A target state is less vulnerable to import restrictions when they have a domestic market for the commodity.

Alternatively, the target state can adapt to living with the gap. One option available to the target state is to locate viable substitutions for the threatened commodity. We can model how the target state would increase its access to these substitutes similarly to how we measure their expected access to the threaten commodity.

> H3a: A target state is less vulnerable to export restrictions when they are able to substitute demand for the commodity with demand for an alternate product to which they have access.
>
> H3b: A target state is less vulnerable to import restrictions when they are able to substitute their production and sales of that commodity for production and sales of a different commodity.

Alternatively, the target state can attempt to reduce its demand for that commodity and any downstream products that rely on access to that commodity. This would result in an overall reduction in the amount the target state consumes of, or generates revenue from, a commodity.

> H4a: A target state is less vulnerable to export restrictions when it can reduce its overall demand for the commodity in its domestic market.
>
> H4b: A target state is less vulnerable to import restrictions when it has slack in its gross domestic product.

If the target state cannot fill the gap or adapt to living with the gap, it is trade dependent on the imposing state. We should expect to be able to observe this dependence through greater policy convergence [@flores-macías2013; @strüver2014] or direct coercion [@kastner2016]. In short, the target state is vulnerable to coercion.

In summary, whether or not target state is trade dependent on an imposing state is a product of the target's ability to fill the gap created or threatened by the imposing state. This is intuitive, but it often poorly measured in international relations scholarship.

## Empirics

### Identifying trade dependence

Can the imposing state credibly threaten to impose severe economic costs on the target state? To answer this question, we need to measure trade market concentration. At the extreme, if the imposing state is the only supplier of a commodity, the target state will not be able to locate alternative suppliers. It will either need to find a substitute for, or reduce its consumption of, the commodity and any downstream products. Alternatively, if the imposing state is one of many suppliers of the commodity, the target state should be able to locate an alternative supplier and only suffer slightly increased prices for the commodity on the global markets.

The Herfindahl-Hirschman Index (HHI) is a widely used measure of market concentration. When applied to international trade markets, the HHI measures the total value of a commodity traded by each state relative to the total value of trade in the commodity. It is calculated as the sum of the square of each state's market share. It ranges from zero to 10,000, with values nearing zero indicating a highly competitive market and values nearing 10,000 indicating a highly concentrated or monopolized market.

This analysis uses the HHI to identify commodity trade markets in which the dominant participant could credibly threaten to strangle a target state's international access to the commodity. I examine market concentration for all commodities traded globally across our period of interest. This includes `r nrow(product_codes) |> comma()` different commodities.[^1] I use global commodity-level trade data provided by UN COMTRADE.

[^1]: This analysis defines commodity groupings according to the four digit Harmonized Tariff Schedule (HS). The HS is a standardized classification system for traded products. It is used globally to identify products when collecting taxes, assessing duties, and reporting trade statistics. A full list of commodities and their HS codes can be accessed from [UN COMTRADE](https://comtrade.un.org/Data/cache/classificationHS.json).

Across these `r nrow(product_codes) |> comma()` different commodities, the HHI ranges from `r hhi_summary |> summarise(min(hhi)) |> round()` to `r hhi_summary |> summarise(max(hhi)) |> round() |> pull() |> comma()`. Commonly, a market with a HHI of more than 2,500 is considered to be highly concentrated. `r hhi_summary |> count(hc = hhi > 2500) |> mutate(prop = n / sum(n)) |> filter(hc == T) |> pull(prop) |> percent()` of commodities are traded in highly concentrated markets for at least one year in our period of interest. These highly concentrated markets represent the trade in commodities for which a state or a small set of states can potentially impose severe economic costs on a target state. The dominant trading states in these markets represent potential viable imposing states. Their trade partners represent potential target states.

```{r}
ggplot(hhi_summary, aes(x = hhi)) + 
  geom_histogram() +
  geom_vline(xintercept = 2500) + 
  theme_bw() + 
  labs(
    x = "HHI",
    y = NULL
  ) + 
  scale_x_continuous(limits = c(0, 10000))
```

We can now examine the target state's response to a potential trade disruption of a vulnerable commodity. First we will determine whether the target state can fill the gap created by the trade ban. The target can either look to other trading states to fill this gap, or it can increase domestic production in the commodity.

Can the target state locate alternative trade partners to fill the gap left by the imposing state? To measure this, I use annual bilateral trade data describing trade between the most dominant trading state and all of their trade partners for all commodities identified to be traded in highly concentrated markets. Attracted by the higher prices exports would receive or lower prices at which imports could be accessed, could alternative trading partners divert their trade in the vulnerable commodity to the target and fill the gap? For `r hv_summary |> distinct(direction, reporter, partner, cmd_code) |> nrow() |> broman::spell_out()` dyad-commodity pairings, the potential imposing state dominates trade to such an extent that all alternative trade partners combined cannot fill the gap they could leave. A full list of these products and trade partners is provided in the appendix. 

For the remaining dyad-commodity pairings, trade diverted from some combination of alternative trade parters could fill the gap left by the imposing state.   

## Measuring influence

We first need an appropriate measure of political influence. It is difficult to operationalize political influence across countries and time [@kastner2016].

Both Flores-Macias and Kreps [-@flores-macías2013] and Struver [-@strüver2014] measure the distance between a country's and China's voting pattern in the United Nations General Assembly. Kastner [-@kastner2016] examines countries' positions on issues pertaining to issues over which China cares deeply: Taiwan and Tibet, and its status as a market economy.

## Appendix

**Complete list of dominated pairings:**

```{r}
hv_summary |> 
  arrange(reporter, partner, cmd_code, year) |> 
  rename(commodity_code = cmd_code, description = cmd_desc) |> 
  janitor::clean_names("title") |>
  knitr::kable()
```
